A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-26 latex/main.tex
Update hte pango X events table

% This table is generated by the notebook tab_pango_x_events
\begin{table}
\caption{
ARG events associated with Pango X lineages.
Each row corresponds to the oldest node assigned the specified
Pango lineage and the ancestor of all nodes assigned that label
(except XBB, XBB.x and XM; see STAR methods). [MORE DETAILS]
}
\label{tab:pango_x_lineages}
\begin{footnotesize}
\begin{tabular}{lrrrllrl}
\toprule
\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
pango & avt & left & right & left p & right p & samp & nested \\
\midrule
XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
XAL:3} \\
XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {} \\
XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
\midrule
\multicolumn{5}{l}{Type II RE event}\\
alias & avt & left & right & left p & right p & desc & by pango \\
\midrule
Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 & {BA.2:156,
XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
{XE:1116, BA.2:37, XH:2, XAF:1} \\
XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
BF.3:2} \\
XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
XAF:1} \\
XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
BA.2:17} \\
\bottomrule
\multicolumn{5}{l}{Events closely derived from RE}\\
pango & alias & path len & time & mutations & samples \\
\midrule
XR & XQ & 3 & 11 & 4 & 17 \\
XU & XQ & 6 & 103 & 13 & 1 \\
XAM & XQ & 6 & 40 & 3 & 21 \\
XAA & XQ & 6 & 33 & 3 & 17 \\
XAG & XQ & 6 & 53 & 4 & 6 \\
XAP & Xx & 4 & 66 & 1 & 20 \\
XZ & Xx & 4 & 48 & 1 & 48 \\
XAE & Xx & 2 & 47 & 7 & 9 \\
XAF & XAF & 5 & 140 & 8 & 1 \\
XAL & XM & 3 & 65 & 2 & 3 \\
\midrule
\multicolumn{5}{l}{Non recombinant}\\
pango & muts & parent & samples & nested \\
\midrule
XP & 1 & \textbf{BA.1.1} & 45 & {} \\
XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
XBQ & 1 & CJ.1 & 14 & {} \\
XBK & 1 & CJ.1 & 7 & {} \\
XBE & 3 & BA.5.2 & 65 & {} \\
XB & 2 & B.1 & 192 & {} \\
XAZ & 3 & \textbf{BA.5} & 133 & {} \\
XAV & 3 & BA.5.1.24 & 13 & {} \\
XAU & 1 & BA.2 & 8 & {} \\
XAS & 1 & BA.4 & 77 & {} \\
XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table}

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed a more comprehensive mutational landscape
of SARS-CoV-2 by mapping deletions onto the ARG
using post-hoc parsimony.
We identified important deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All these deletions are known to occur major VOCs
(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
\cite{Akaishi2023,Li2023indels}).
Eight of these deletions occur in S1-NTD,
a known hotspot for recurrent deletions \cite{McCarthy2021}.
""" The 5-base dels and 4-base del are split up as separate events.
It should be a 9-base del (11,288–11,296)."""
Furthermore, these deletion events reflect
the VOC-associated deletion events
inferred in a previous study \cite{Akaishi2023}.

>>>>

\begin{table}
\caption{
ARG events associated with Pango X lineages.
Each row for Type I, III and IV events corresponds to the oldest node 
assigned the specified Pango lineage and is the ancestor of all 
nodes assigned that label. Type II events are recombination events
associated with Pango X lineages not meeting Type I criteria.
For Type I and II recombination 
events, the averted shows the number of additional mutations required
under a recombination-free HMM solution. Parent lineages shown in boldface
are nodes exactly matched by sampled sequences. The descendants column
shows the descendant samples broken down by Pango lineage (grouped 
by root lineage, where appropriate).
}
\label{tab:pango_x_lineages}
% scriptsize is quite small here, and we could probably cram it all 
% in with footnotesize if we worked a bit harder on the intercol
% spacing
\begin{scriptsize}
\begin{tabular}{lrrlrll}
\toprule
\multicolumn{7}{l}{Type I event: Recombination coinciding with Pango origin node}\\
&&\multicolumn{2}{c}{Parent lineage} & \multicolumn{2}{c}{Interval}\\
pango & averted & left & right & left & right & descendants \\
\midrule

XC & 37 & \textbf{AY.29} & \textbf{B.1.1.7} & 26768 & 27390 & {XC:5} \\
XBR & 22 & \textbf{BN.3.1} & BQ.1.25 & 22034 & 22190 & {XBR:1} \\
XA & 16 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 20411 & 21765 & {XA:39} \\
XS & 11 & AY.103 & \textbf{BA.1.1} & 9054 & 10449 & {XS:17} \\
XL & 8 & \textbf{BA.1.17.2} & \textbf{BA.2.5} & 5925 & 8393 & {XL:64} \\
XBG & 8 & \textbf{BA.2.76} & \textbf{BA.5.2} & 22305 & 22917 & {XBG:25} \\
XQ & 7 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 4322 & 5386 & {XQ:55, BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBD & 7 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 23020 & 24620 & {XBD:30} \\
XM & 6 & \textbf{BA.1.1} & \textbf{BA.2} & 17411 & 21595 & {XM:26, BA.2:16,
XAL:3} \\
XBH & 6 & \textbf{BA.2.1} & BA.2.75.2 & 15452 & 22001 & {XBH:6} \\
XBB & 6 & \textbf{BA.2.10} & BM.1.1.1 & 22332 & 22577 & {XBB:6452, BA.2:1} \\
XY & 5 & \textbf{BA.1.1} & \textbf{BA.2} & 11538 & 12880 & {XY:23} \\
XF & 5 & \textbf{AY.4} & \textbf{BA.1} & 5387 & 6402 & {XF:16} \\
XBF & 5 & \textbf{BA.5.2.1} & CJ.1 & 5184 & 9866 & {XBF:185} \\
XW & 4 & \textbf{BA.1.1.15} & \textbf{BA.2} & 2833 & 4321 & {XW:32} \\
XG & 4 & \textbf{BA.1.17} & \textbf{BA.2} & 5925 & 6513 & {XG:3} \\

\midrule
\multicolumn{7}{l}{Type II event: Recombination closely associated with pango lineage(s)}\\
&&\multicolumn{2}{c}{Parent lineage} & \multicolumn{2}{c}{Interval}\\
name & averted & left & right & left & right & descendants \\
\midrule

Xx & 6 & \textbf{BA.2} & \textbf{BA.1.17.2} & 24504 & 26060 & {BA.2:156, XZ:48,
XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & \textbf{BA.1.17.2} & \textbf{BA.2} & 10448 & 11283 & {XE:1116,
BA.2:37, XH:2, XAF:1} \\
XBM & 6 & \textbf{BA.2.76} & \textbf{BF.3} & 22305 & 22917 & {XBM:10, BF.3:2}
\\
XJ & 5 & \textbf{BA.1.17.2} & \textbf{BA.2} & 13196 & 17410 & {XJ:68, BA.2:17}
\\
XAF & 5 & \textbf{BA.1.1} & \textbf{BA.2} & 10199 & 10447 & {BA.2:35, XAF:1} \\

\midrule
\multicolumn{7}{l}{Type III events: Pango origin nodes derived from 
Type I and II events}\\
pango & name & path len & time & mutations & samples \\
\midrule

XAF & XAF & 5 & 140 & 8   && {XAF:1} \\
XBM & XBM & 1 & 23 & 1    && {XBM:10} \\
XE  & XE/XH & 1 & 19 & 2  && {XE:1116} \\
XH  & XE/XH & 1 & 55 & 7  && {XH:2} \\
XJ  & XJ & 1 & 5 & 2      && {XJ:68, BA.2:3} \\
XAL & XM & 3 & 65 & 2     && {XAL:3} \\
XAA & XQ & 6 & 33 & 3     && {XAA:17} \\
XAG & XQ & 6 & 53 & 4     && {XAG:6} \\
XAM & XQ & 6 & 40 & 3     && {XAM:21} \\
XR  & XQ & 3 & 11 & 4     && {XR:17, BA.2:1} \\
XU  & XQ & 6 & 103 & 13   && {XU:1} \\
XAE & Xx & 2 & 47 & 7     && {XAE:9} \\
XAP & Xx & 4 & 66 & 1     && {XAP:20} \\
XZ  & Xx & 4 & 48 & 1     && {XZ:48} \\

\midrule
\multicolumn{7}{l}{Type IV events: non recombinants}\\
pango & muts & parent & samples & nested \\
\midrule

XP  & 1  &&&& \textbf{BA.1.1} & {XP:45} \\
XN  & 1  &&&& \textbf{BA.2} & {XN:120, BA.2:1} \\
XBQ & 1  &&&& CJ.1 & {XBQ:14} \\
XBK & 1  &&&& CJ.1 & {XBK:7} \\
XBE & 3  &&&& BA.5.2 & {XBE:65} \\
XB  & 2  &&&& B.1 & {XB:192} \\
XAZ & 3  &&&& \textbf{BA.5} & {XAZ:133} \\
XAV & 3  &&&& BA.5.1.24 & {XAV:13} \\
XAU & 1  &&&& BA.2 & {XAU:8} \\
XAS & 1  &&&& BA.4 & {XAS:77} \\
XAN & 5  &&&& \textbf{BA.5.1} & {XAN:7} \\
XAJ & 11 &&&& \textbf{BA.2.12} & {XAJ:18} \\

\bottomrule
\end{tabular}
\end{scriptsize}
\end{table}

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed a more comprehensive mutational landscape
of SARS-CoV-2 by mapping deletions onto the ARG
using post-hoc parsimony.
We identified important deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All these deletions are known to occur major VOCs
(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
\cite{Akaishi2023,Li2023indels}).
Eight of these deletions occur in S1-NTD,
a known hotspot for recurrent deletions \cite{McCarthy2021}.
There are three contiguous deletions in NSP6
(4-base at 11288, 5-base at 11283, and 5-base at 11292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node and
the other above the BA.2 origin node in the ARG
(Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up
during deletion remapping because they had
overlapping four sites at 11288-11292
(which was mapped as the 4-base deletion event).
Furthermore, the placement of these deletion events in the ARG
reflects that of the VOC-associated deletion events
inferred in a previous study \cite{Akaishi2023}.
